Canadian Grocery Price Analysis

### 2024-12-04 paper/paper.qmd
Add files via upload

# Load model
pricing_model <- readRDS(file = here("models", "pricing_model.rds"))
```

@@

1. **Data Preparation**: Categorical variables (e.g., retailer name, month, product type) are encoded as factors to enable straightforward interpretation of the model coefficients. Continuous variables are scaled where necessary to ensure numerical stability during model fitting.
2. **MCMC Sampling**: Markov Chain Monte Carlo (MCMC) is used to draw samples from the posterior distributions of the parameters. Diagnostic measures, such as trace plots and \( \hat{R} \) values, are used to verify convergence and ensure reliable results.
3. **Posterior Predictive Checks**: Simulated datasets are generated from the posterior distribution to compare predicted values with observed data, ensuring that the model captures key features of the pricing patterns.

@@

The following diagnostics are performed to assess the model's validity:

- **Convergence Checks**: Trace plots and \( \hat{R} \) values are examined to confirm that the MCMC chains have converged and are exploring the posterior distribution effectively.
- **Posterior Predictive Checks**: Visualizations of observed versus simulated data are used to evaluate the model's ability to replicate the data's structure.
- **Residual Analysis**: Residuals are plotted and examined for patterns or deviations, verifying the linearity and homoscedasticity assumptions.

@@

### Cleaning Steps
The following transformations were applied to prepare the dataset for analysis:
- **Imputation of missing values**:
  - Missing prices (`current_price`, `old_price`) were replaced with vendor-specific averages.
  - Missing `price_per_unit` values were computed using:
    \[
    \text{price\_per\_unit} = \frac{\text{current\_price}}{\text{units}}
    \]
- **Normalization**:
  - Prices were converted to Canadian dollars (CAD) if discrepancies were noted.
  - `current_price` and `old_price` were standardized using `parse_number()` to handle non-numeric characters.
- **Outlier removal**:
  - Prices exceeding three times the interquartile range (IQR) were flagged and removed.
- **Variable creation**:
  - Derived variables included `price_difference` (old price minus current price) and `month` (from timestamps).

@tbl-cleaned_data showcases the cleaned dataset.

```{r}
#| label: tbl-cleaned_data
#| tbl-cap: "Summary of Cleaned Dataset"
#| echo: false

cleaned_data_summary <- tibble(
  Variable = c("current_price", "old_price", "vendor", "price_per_unit", "month"),
  Type = c("Numeric", "Numeric", "Categorical", "Numeric", "Categorical"),
  Description = c(
    "Current retail price of beef (CAD)",
    "Historical price of beef (CAD)",
    "Retailer identifier (Walmart, T&T)",
    "Price per unit of beef (CAD per unit)",
    "Month extracted from transaction timestamp"
  )
)

cleaned_data_summary %>%
  kable(col.names = c("Variable", "Type", "Description"), booktabs = TRUE)
```

## Model Diagnostics {#sec-model-diagnostics}

pp_check(beef_model, plotfun = "dens_overlay")
```

plot(beef_model, plotfun = "trace")
```

plot(beef_model, plotfun = "rhat")
```

### Model Fit Metrics
The model achieved a high adjusted \(R^2\) of 0.94, indicating strong predictive power. Additional diagnostics include:
- **WAIC**: Widely Applicable Information Criterion for model comparison.
- **LOOIC**: Leave-One-Out Cross-Validation Information Criterion.
- **RMSE**: Root Mean Square Error to measure prediction accuracy.

model_diagnostics_summary <- tibble(
  Metric = c("Adjusted R-Squared", "WAIC", "LOOIC", "RMSE"),
  Value = c(0.94, 8709, 8707, 1.02)
)

### Cleaning Steps
The following transformations were applied to prepare the dataset for analysis:
- **Imputation of missing values**:
  - Missing prices (`current_price`, `old_price`) were replaced with vendor-specific averages.
  - Missing `price_per_unit` values were computed using:
    \[
    \text{price\_per\_unit} = \frac{\text{current\_price}}{\text{units}}
    \]
- **Normalization**:
  - Prices were converted to Canadian dollars (CAD) if discrepancies were noted.
  - `current_price` and `old_price` were standardized using `parse_number()` to handle non-numeric characters.
- **Outlier removal**:
  - Prices exceeding three times the interquartile range (IQR) were flagged and removed.
- **Variable creation**:
  - Derived variables included `price_difference` (old price minus current price) and `month` (from timestamps).

@tbl-cleaned_data_summary showcases the cleaned dataset.

```{r}
#| label: tbl-cleaned_data_summary
#| tbl-cap: "Summary of Cleaned Dataset"
#| echo: false

cleaned_data_summary <- tibble(
  Variable = c("current_price", "old_price", "vendor", "price_per_unit", "month"),
  Type = c("Numeric", "Numeric", "Categorical", "Numeric", "Categorical"),
  Description = c(
    "Current retail price of beef (CAD)",
    "Historical price of beef (CAD)",
    "Retailer identifier (Walmart, T&T)",
    "Price per unit of beef (CAD per unit)",
    "Month extracted from transaction timestamp"
  )
)

cleaned_data_summary %>%
  kable(col.names = c("Variable", "Type", "Description"), booktabs = TRUE)
```

## Model Diagnostics {#sec-model-diagnostics}

pp_check(beef_model, plotfun = "dens_overlay")
```

plot(beef_model, plotfun = "trace")
```

plot(beef_model, plotfun = "rhat")
```

### Model Fit Metrics
The model achieved a high adjusted \(R^2\) of 0.94, indicating strong predictive power. Additional diagnostics include:
- **WAIC**: Widely Applicable Information Criterion for model comparison.
- **LOOIC**: Leave-One-Out Cross-Validation Information Criterion.
- **RMSE**: Root Mean Square Error to measure prediction accuracy.

model_diagnostics_summary <- tibble(
  Metric = c("Adjusted R-Squared", "WAIC", "LOOIC", "RMSE"),
  Value = c(0.94, 8709, 8707, 1.02)
)

>>>>

# Load model
pricing_model <- readRDS(file = here("models", "model.rds"))
```

@@

1. **Data Preparation**: Categorical variables (e.g., retailer name, month, product type) are encoded as factors to enable straightforward interpretation of the model coefficients. Continuous variables are scaled where necessary to ensure numerical stability during model fitting.
2. **MCMC Sampling**: Markov Chain Monte Carlo (MCMC) is used to draw samples from the posterior distributions of the parameters. Diagnostic measures, such as trace plots and $$ \hat{R} $$ values, are used to verify convergence and ensure reliable results.
3. **Posterior Predictive Checks**: Simulated datasets are generated from the posterior distribution to compare predicted values with observed data, ensuring that the model captures key features of the pricing patterns.

@@

pp_check(pricing_model, plotfun = "dens_overlay")
```

plot(pricing_model, plotfun = "trace")
```

plot(pricing_model, plotfun = "rhat")
```

### Model Fit Metrics
The model achieved a high adjusted \(R^2\), indicating strong predictive power. Additional diagnostics include:
- **WAIC**: Widely Applicable Information Criterion for model comparison.
- **LOOIC**: Leave-One-Out Cross-Validation Information Criterion.
- **RMSE**: Root Mean Square Error to measure prediction accuracy.

model_diagnostics_summary <- tibble(
  Metric = c("Adjusted R-Squared", "WAIC", "LOOIC", "RMSE"),
  Value = c(NA, NA, NA, NA) # Replace NA with actual values after computation
)

@@
